---
title: "Preparing Spatial Data"
author: "Divyashree Rana"
date: "`r Sys.Date()`"

output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dpi = 300
)

```

# Load required packages

The following package is required for raster processing, moving window analyses, and exporting resistance surfaces for subsequent optimization.

```{r load_packages}

library(terra)

```

# Load 1 km raster surfaces

Environmental predictor rasters are first imported at their native 1 km resolution from Google Earth Engine and stored in ./input/rasters_1km. These rasters serve as the baseline inputs from which coarser-scale representations are generated using moving window smoothing. This allows rasters processed at different spatial scales to have same resolution and still mergeable for multisurface optimization. 


## Import raster layers

```{r load_rasters}

raster_dir <- "./input/rasters_1km"

files <- list.files(
  raster_dir,
  pattern = "\\.asc$",
  full.names = TRUE
)

files

rasters <- rast(files)

rasters

```

## Check raster consistency

All raster layers should have identical extent, resolution, and coordinate reference system before moving window analyses are performed.

```{r check_geometry}

compareGeom(
  rasters[[1]],
  rasters[[2]]
)

compareGeom(
  rasters[[1]],
  rasters[[3]]
)

compareGeom(
  rasters[[1]],
  rasters[[4]]
)
```

# Generate smoothed raster surfaces

Moving window analyses are used to summarize environmental conditions across broader neighbourhoods while maintaining the original 1 km raster resolution. Circular moving windows with radii of 2 km, 5 km, and 10 km are applied using focal means.

## Define moving windows

```{r moving_windows}

w2 <- focalMat(
  rasters[[1]],
  d = 2000,
  type = "circle"
)

w5 <- focalMat(
  rasters[[1]],
  d = 5000,
  type = "circle"
)

w10 <- focalMat(
  rasters[[1]],
  d = 10000,
  type = "circle"
)
```

## Applying moving window smoothening to the 1km rasters

```{r applying_moving_windows}
rasters_2km <- focal(
  rasters,
  w = w2,
  fun = mean,
  na.rm = TRUE
)

rasters_5km <- focal(
  rasters,
  w = w5,
  fun = mean,
  na.rm = TRUE
)

rasters_10km <- focal(
  rasters,
  w = w10,
  fun = mean,
  na.rm = TRUE
)
```

## Verify smoothed rasters

Moving window smoothing should preserve raster geometry and should not introduce missing values along raster edges.

```{r verify_rasters}

compareGeom(
  rasters,
  rasters_2km
)

compareGeom(
  rasters,
  rasters_5km
)

compareGeom(
  rasters,
  rasters_10km
)

global(
  rasters_2km,
  fun = function(x) sum(is.na(x))
)

global(
  rasters_5km,
  fun = function(x) sum(is.na(x))
)

global(
  rasters_10km,
  fun = function(x) sum(is.na(x))
)

```
## Visualize smoothed raster surfaces

```{r plot_smoothed_rasters, fig.width=8, fig.height=8}

par(mfrow = c(2,2), mar = c(2,2,2,4))

plot(
  rasters,
  main = paste(names(rasters), "\n1 km")
)

plot(
  rasters_2km,
  main = paste(names(rasters_2km), "\n2 km")
)

plot(
  rasters_5km,
  main = paste(names(rasters_5km), "\n5 km")
)

plot(
  rasters_10km,
  main = paste(names(rasters_10km), "\n10 km")
)

par(mfrow = c(1,1))

```

# Export raster surfaces

Raster stacks from each spatial scale are exported as ASCII grids for use in subsequent resistance surface optimization analyses. Create a ./input/smooth folder for storing these processed layers of at multiple scales all with a resolution of 1km.

```{r export_rasters}

raster_list <- list(
  "1km" = rasters,
  "2km" = rasters_2km,
  "5km" = rasters_5km,
  "10km" = rasters_10km
)

base_dir <- "./input/smooth"

for(scale_name in names(raster_list)) {

  r_stack <- raster_list[[scale_name]]

  out_dir <- file.path(
    base_dir,
    paste0("rasters_", scale_name)
  )
  
  ## Create output directory if it does not already exist
  if(!dir.exists(out_dir)){

    dir.create(
      out_dir,
      recursive = TRUE,
      showWarnings = FALSE
    )

  }

  for(i in 1:nlyr(r_stack)) {

    writeRaster(

      r_stack[[i]],

      filename = file.path(
        out_dir,
        paste0(
          names(r_stack)[i],
          "_",
          scale_name,
          ".asc"
        )
      ),

      overwrite = TRUE,

      filetype = "AAIGrid"

    )

  }

}
```


